---
title: "HFCE Food Time Series Using Seasonal ARIMA"
author: "<font size=\"2\"> Christony G. Duyapat ^[Saint Mary's University - School of Graduate Studies, hed-cduyapat@smu.edu.ph]</font>"
output:
  html_document:
    self_contained: yes
    mode: selfcontained
---

------------------------------------------------------------------------

The data used in this report were downloaded from [Philippines Statistics Authority](https://psa.gov.ph/statistics/national-accounts/data-series) website accessed on **April 6, 2024**.

### Six Major Steps in Developing a Forecast

1.  Defining the Problem

2.  Gathering and Preprocessing Data

3.  Exploratory Analysis

4.  Choosing and Building Models

5.  Computing Model Performance Metrics and Checking Residuals

6.  Choosing the Best Model and Making Forecast

------------------------------------------------------------------------

#### Step 1: Defining the Problem

Research Questions:

1.  What is the spending pattern of Filipinos on food for the first quarter of 2000 to fourth quarter of 2023?

2.  What is the best ARIMA model in forecasting the Household Final Consumption Expenditure on Food?

3.  What are the forecasted values of the Household Final Consumption Expenditure on Food for the years 2024 to 2025?

------------------------------------------------------------------------

#### Step 2: Gathering and Preprocessing Data

Load the **library packages** to be used in the analysis.

```{r}
knitr::opts_chunk$set(dev='svg')
library(forecast)
library(ggplot2)
library(caret)
```

Import the data and transform it into a time series data.

```{r}
data=read.csv("hfce_food.csv")
food_ts=ts(data$food,
           start=c(2000,1), 
           end=c(2023,4),
           frequency=4)
print(food_ts)
summary(food_ts)
```

------------------------------------------------------------------------

#### Step 3: Exploratory Analysis

Plot the time series data.

```{r}
autoplot(food_ts,
         main='HFCE on Food 2000-2023', 
         xlab='Quarter',
         ylab='in million Php')
```

Time Series Decomposition and Seasonal Plot

```{r}
autoplot(decompose(food_ts))
ggseasonplot(food_ts,main="Seasonal Plot",ylab="in million Php")
```

Research Question 1: What is the spending behavior of Filipinos on food for the first quarter of 2000 to fourth quarter of 2023?

Over the past 24 years, the Household Final Consumption Expenditure (HFCE) on food has exhibited an upward trend with remarkable seasonal patterns. Specifically, there is a spike in spending every second quarter, followed by a decrease in the third quarter. Notably, there is a major increase every fourth quarter which is likely due to increased social gatherings and Christmas-related events that involve higher food consumption. Additionally, the second and fourth-quarter spending spikes may also be partly attributed to the distribution of midyear and Christmas bonuses to employees, respectively.

We can use the Autocorrelation Function and Partial Autocorrelation Function if we want to manually select the model. In my case, I am going to use the automated model selection under the forecast package. Nevertheless, let us show the ACF and PACF for demonstration.

```{r}
acf(food_ts,lag.max=36)
pacf(food_ts,lag.max=36)
```

------------------------------------------------------------------------

#### Step 4: Building and Choosing Models

Automated model selection using auto.arima() function.

```{r}
auto.arima(food_ts,trace=TRUE,ic="bic",stepwise=FALSE)
```

Choose top 5 best models using Bayesian Information Criterion (lower the better) and build the ARIMA models.

1.  ARIMA(0,1,1)(0,1,0)[4] : 2030.392

2.  ARIMA(1,1,0)(0,1,0)[4] : 2031.81

3.  ARIMA(1,1,1)(0,1,0)[4] : 2034.512

4.  ARIMA(0,1,1)(1,1,0)[4] : 2034.615

5.  ARIMA(0,1,1)(0,1,1)[4] : 2034.701

```{r}
model1=arima(food_ts,order=c(0,1,1),seasonal=list(order=c(0,1,0)))
model2=arima(food_ts,order=c(1,1,0),seasonal=list(order=c(0,1,0)))
model3=arima(food_ts,order=c(1,1,1),seasonal=list(order=c(0,1,0)))
model4=arima(food_ts,order=c(0,1,1),seasonal=list(order=c(1,1,0)))
model5=arima(food_ts,order=c(0,1,1),seasonal=list(order=c(0,1,1)))
```

Compare the Bayesian Information Criteria and Akaike Information Criteria.

```{r}
rbind(Model1=data.frame(BIC=BIC(model1),AIC=AIC(model1)),
      Model2=data.frame(BIC=BIC(model2),AIC=AIC(model2)),
      Model3=data.frame(BIC=BIC(model3),AIC=AIC(model3)),
      Model4=data.frame(BIC=BIC(model4),AIC=AIC(model4)),
      Model5=data.frame(BIC=BIC(model5),AIC=AIC(model5)))
```

Visualize the top 5 models on top of the actual values

```{r}
model_fit1=fitted.values(model1)
model_fit2=fitted.values(model2)
model_fit3=fitted.values(model3)
model_fit4=fitted.values(model4)
model_fit5=fitted.values(model5)

autoplot(food_ts,series="Actual Values",main="Model 1 Fitted Values")+
  autolayer(model_fit1,series="ARIMA(0,1,1)(0,1,0)[4]")+
  xlab("Quarter")+
  ylab("in million Php")+
  guides(colour=guide_legend(title=""))+
  scale_color_manual(values=c('black','aquamarine3'))

autoplot(food_ts,series="Actual Values",main="Model 2 Fitted Values")+
  autolayer(model_fit2,series="ARIMA(1,1,0)(0,1,0)[4]")+
  xlab("Quarter")+
  ylab("in million Php")+
  guides(colour=guide_legend(title=""))+
  scale_color_manual(values=c('black','cadetblue3'))

autoplot(food_ts,series="Actual Values",main="Model 3 Fitted Values")+
  autolayer(model_fit3,series="ARIMA(1,1,1)(0,1,0)[4]")+
  xlab("Quarter")+
  ylab("in million Php")+
  guides(colour=guide_legend(title=""))+
  scale_color_manual(values=c('black','chartreuse2'))

autoplot(food_ts,series="Actual Values",main="Model 4 Fitted Values")+
  autolayer(model_fit4,series="ARIMA(0,1,1)(1,1,0)[4]")+
  xlab("Quarter")+
  ylab("in million Php")+
  guides(colour=guide_legend(title=""))+
  scale_color_manual(values=c('black','darkgoldenrod2'))

autoplot(food_ts,series="Actual Values",main="Model 5 Fitted Values")+
  autolayer(model_fit5,series="ARIMA(0,1,1)(0,1,1)[4]")+
  xlab("Quarter")+
  ylab("in million Php")+
  guides(colour=guide_legend(title=""))+
  scale_color_manual(values=c('black','darkorchid2'))
```

------------------------------------------------------------------------

#### Step 5: Computing Model Performance Metrics and Checking Residuals

Compute the Performance Metrics of the Models

```{r}
rbind(
  Model1_PerfMetrics=data.frame(accuracy(model1),R2=R2(model_fit1,food_ts)),
  Model2_PerfMetrics=data.frame(accuracy(model2),R2=R2(model_fit2,food_ts)),
  Model3_PerfMetrics=data.frame(accuracy(model3),R2=R2(model_fit3,food_ts)),
  Model4_PerfMetrics=data.frame(accuracy(model4),R2=R2(model_fit4,food_ts)),
  Model5_PerfMetrics=data.frame(accuracy(model5),R2=R2(model_fit5,food_ts))
  )

```

Check the residuals if they are independently distributed using Ljung-Box Test.

```{r}
rbind(Model1_Ljung=checkresiduals(model1),
      Model2_Ljung=checkresiduals(model2),
      Model3_Ljung=checkresiduals(model3),
      Model4_Ljung=checkresiduals(model4),
      Model5_Ljung=checkresiduals(model5)
      )
```

The Ljung-Box test shows that only Model 2 (**ARIMA(1,1,0)(0,1,0)[4]**) accepts the null hypothesis at *p-value=0.1623535*. Ljung-Box test is used to show that the residuals are independently distributed which is one of the assumptions of ARIMA.

------------------------------------------------------------------------

#### Step 6: Choosing the Best Model and Making Forecast

Research Question 2: What is the best ARIMA model in forecasting the Household Final Consumption Expenditure on Food?

Upon considering the performance metrics and residuals, Model 2 **ARIMA(1,1,0)(0,1,0)[4]** is evaluated as the best model.

This model is represented by the equation: y~t~ = 𝜙~1~y~t-1~ - 𝜙~1~y~t-5~ - 𝜙~1~y~t-2~ + 𝜙~1~y~t-6~ + y~t-4~ + y~t-1~ - y~t-5~ + 𝜀~t~

```{r}
summary(model2)
```

Research Question 3: What are the forecasted values of the Household Final Consumption Expenditure on Food for the years 2024 to 2025?

Make forecast using the best model for the next 8 quarters or in the next 2 years.

```{r}
forecast=forecast(model2,h=8)
autoplot(food_ts,series="Actual Values",main="Forecasts from ARIMA(1,1,0)(0,1,0)[4]")+
  autolayer(forecast,series="Confidence Interval")+
  autolayer(forecast$mean,series="Forecasted Values")+
  xlab("Quarter")+
  ylab("in million Php")+
  guides(colour=guide_legend(title=""))+
  scale_color_manual(values=c('black','darkorchid4','blue'))
options(digits=12)
print(forecast(model2,h=8))
```

